---
title: "cmml/cord biological reps simulation"
author: "Christian La France"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


This markdown simulates biological replicates from the original cmml/cord data, from a normal distribution, and from a poisson distribution. The normal and poisson distributions were generated using the mean pseudotime values from the real cmml and cord data sets and have the same number of values as the real data sets. 


```{r}
library(tidyverse)
library(Signac)
library(SeuratObject)
library(ggplot2)
library(nortest)
```


```{r}
query_object_with_PT = readRDS("C:/Users/chris/Downloads/cord-cmml-granja-mapped-with-median-PT_no_clusters_0_13.rds")
```


```{r}
# fetch relevant data
cord_cmml_medians <- FetchData(query_object_with_PT, vars = c("orig.ident", "median_pseudotime")) %>% group_by(orig.ident) %>% as.data.frame()

ggplot() + geom_density(data=cord_cmml_medians, aes(x=median_pseudotime, group=orig.ident, fill=orig.ident), alpha=0.5) + labs(title="Density of pseudotime values")
```


```{r}
# separate cord and cmml
cord_PTs <- cord_cmml_medians[cord_cmml_medians$orig.ident=="cord", ]
cmml_PTs <- cord_cmml_medians[cord_cmml_medians$orig.ident=="cmml", ]
```


```{r}
# set parameters for sampling from distributions
n_cord = nrow(cord_PTs)
n_cmml = nrow(cmml_PTs)

cord_mean_pt <- mean(cord_PTs$median_pseudotime, rm.na=TRUE)
cmml_mean_pt <- mean(cmml_PTs$median_pseudotime, rm.na=TRUE)

cord_pt_sd <- sd(cord_PTs$median_pseudotime)
cmml_pt_sd <- sd(cmml_PTs$median_pseudotime)

n = 5 # number of biological replicates
```


### Simulate biological replicates from original data (probably not as useful?).
```{r}
n_adjust = 0.8 # adjust sample size to ensure variation. Only used for simulation from real data. 


cord_real_sim <- data.frame(sample1 = sample(x=cord_PTs$median_pseudotime, size=n_cord*n_adjust, replace=TRUE))
cmml_real_sim <- data.frame(sample1= sample(x=cmml_PTs$median_pseudotime, size=n_cmml*n_adjust, replace=TRUE))

for (i in 2:n) {
  col_name <- paste("sample", i, sep="")
  cord_real_sim[col_name] <- sample(x=cord_PTs$median_pseudotime, size=n_cord*n_adjust, replace=TRUE)
  cmml_real_sim[col_name] <- sample(x=cmml_PTs$median_pseudotime, size=n_cmml*n_adjust, replace=TRUE)
}

cord_real_sim <- pivot_longer(cord_real_sim, 
                                 cols=c("sample1", "sample2", "sample3", "sample4", "sample5"), 
                                 names_to = "sample",
                                 values_to="pseudotime")

cmml_real_sim <- pivot_longer(cmml_real_sim, 
                                 cols=c("sample1", "sample2", "sample3", "sample4", "sample5"), 
                                 names_to = "sample",
                                 values_to="pseudotime")


ggplot(data=cord_real_sim, aes(x=pseudotime, group=sample)) + geom_density() + 
  labs(title=paste("Simulated Cord pseudotime values from real data,", n, "biological reps")) + 
  xlab("Pseudotime values")


ggplot(data=cmml_real_sim, aes(x=pseudotime, group=sample)) + geom_density() + 
  labs(title=paste("Simulated CMML pseudotime values from real data,", n, "biological reps")) +
  xlab("Pseudotime values")

```




### Simulate biological reps from normal distribution
```{r}
# this will include some negative pseudotime values. We may need to find a way
# around this. Cannot use abs() because this would skew away from a normal
# distribution. 

cord_rand_normal <- data.frame(sample1 = rnorm(n=n_cord, mean=cord_mean_pt, sd=cord_pt_sd))
cmml_rand_normal <- data.frame(sample1= rnorm(n=n_cmml, mean=cmml_mean_pt, sd=cmml_pt_sd))

for (i in 2:n) {
  col_name <- paste("sample", i, sep="")
  cord_rand_normal[col_name] <- rnorm(n=n_cord, mean=cord_mean_pt, sd=cord_pt_sd)
  cmml_rand_normal[col_name] <- rnorm(n=n_cmml, mean=cmml_mean_pt, sd=cmml_pt_sd)
}

cord_rand_normal <- pivot_longer(cord_rand_normal, 
                                 cols=c("sample1", "sample2", "sample3", "sample4", "sample5"), 
                                 names_to = "sample",
                                 values_to="pseudotime")

cmml_rand_normal <- pivot_longer(cmml_rand_normal, 
                                 cols=c("sample1", "sample2", "sample3", "sample4", "sample5"), 
                                 names_to = "sample",
                                 values_to="pseudotime")


ggplot(data=cord_rand_normal, aes(x=pseudotime, group=sample)) + geom_density() + 
  labs(title=paste("Simulated Cord pseudotime values from normal distribution,", n, "biological reps")) + 
  xlab("Pseudotime values")


ggplot(data=cmml_rand_normal, aes(x=pseudotime, group=sample)) + geom_density() + 
  labs(title=paste("Simulated CMML pseudotime values from normal distribution,", n, "biological reps")) +
  xlab("Pseudotime values")

```



### Simulate biological reps form poisson distribution
```{r}
# cord_rand_poissons <- rpois(n = nrow(cord_PTs), lambda = mean(cord_PTs$median_pseudotime))
# cmml_rand_poissons <- rpois(n = nrow(cmml_PTs), lambda = mean(cmml_PTs$median_pseudotime))

cord_rand_pois <- data.frame(sample1 = rpois(n=n_cord, lambda=cord_mean_pt))
cmml_rand_pois <- data.frame(sample1= rpois(n=n_cmml, lambda=cmml_mean_pt))

for (i in 2:n) {
  col_name <- paste("sample", i, sep="")
  cord_rand_pois[col_name] <- rpois(n=n_cord, lambda=cord_mean_pt)
  cmml_rand_pois[col_name] <- rpois(n=n_cmml, lambda=cmml_mean_pt)
}

cord_rand_pois <- pivot_longer(cord_rand_pois, 
                                 cols=c("sample1", "sample2", "sample3", "sample4", "sample5"), 
                                 names_to = "sample",
                                 values_to="pseudotime")

cmml_rand_pois <- pivot_longer(cmml_rand_pois, 
                                 cols=c("sample1", "sample2", "sample3", "sample4", "sample5"), 
                                 names_to = "sample",
                                 values_to="pseudotime")


ggplot(data=cord_rand_pois, aes(x=pseudotime, group=sample)) + geom_density() + 
  labs(title=paste("Simulated Cord pseudotime values from poisson distribution,", n, "biological reps")) + 
  xlab("Pseudotime values")


ggplot(data=cmml_rand_pois, aes(x=pseudotime, group=sample)) + geom_density() + 
  labs(title=paste("Simulated CMML pseudotime values from poisson distribution,", n, "biological reps")) +
  xlab("Pseudotime values")

```



### These are the dataframes with the simulated data. 

```{r}
# Simulated from original data
cord_real_sim
cmml_real_sim

# Simulated from normal distribution
cord_rand_normal
cmml_rand_normal

# Simulated from poisson distribution
cord_rand_pois
cmml_rand_pois

```